---
title: "A12: Control Type"
author: "TYPE YOUR NAME HERE"
date: "Due Date: 11:59pm, Mar 27"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Individual Homework

- Upload your Shiny app on [ShinyApps.io](https://www.shinyapps.io).

- Submit your `.Rmd` AND your URL (https://XXXXX.shinyapps.io/XXXXX) on Collab.

- You need to modify the given interactive plot in the following way:

    - Change the control type for "Number of bins" from `selectInput()` to `sliderInput()`. 
    
    - Change the control type for "Bandwidth adjustment" from `sliderInput()` to `selectInput()`.     
    
    - You need to change `inputId` names in `inputPanel()` and some parts inside `renderPlot()` to run Shiny App.


## Individual Homework (Cont'd)

- Don't delete the given interactive plot.

- If you like, you can change the output type from `ioslides_presentation` to `slidy_presentation`.  

- In your submission file, you should include either of the following statements:

   - "I pledge that I have neither given nor received help on this assignment."
   - "I have [received/given] help on this assignment [to/from] [student]; the nature of this support was [elaborate]."

-  Example: https://youmi.shinyapps.io/A12example/

## Given interactive Plot

```{r}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = seq(10, 30, 5), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(iris$Sepal.Length, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Sepal Length", main = "Histogram and Density of Sepal Length from Iris Data")
  
  dens <- density(iris$Sepal.Length, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Modified interactive Plot

```{r}
# Add your codes
```
