---
title: "DS2003: Introduction to R"
author: "Natalie Kupperman & Bruce Corliss"
class #: "2"
date: "8/25/2022"
output:
     html_document:
          toc: TRUE
          toc_depth: 4
          toc_float: true
          toc_collapsed: true
          theme: journal
          code_folding: show
editor_options: 
  chunk_output_type: console
---

<!--- Change font size for headers --->
<style>
h1.title {
  font-size: 30px;
}
h1 {
  font-size: 26px;
}
h2 {
  font-size: 22px;
}
h3 { 
  font-size: 18px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA)
knitr::opts_knit$set(global.par=TRUE, html.table.attributes = '')
```

```{r, include=FALSE}
par(mar=c(2.1,2.1,.5,1.5), mgp=c(1.3,.3,0), tck=-0.01)
```


## 1. R as a Simple Calculator

### Simple calculation

```{r}
1 + 1 #to run:for mac, command + enter; for windows, 

sqrt(36) / 3 + 2 * .5  # 2 + 1

sqrt(36)/3+2*.5

sqrt(36) / (3 + 2 * .5) # remember PEMDAS

```

###  Assigning values to an R object 

```{r cal2}
x <- sqrt(36) / 3 + 2 * .5   # you can also use "=" instead of "<-"
x                            # print content of R object "x" on the screen
```

### Simple computations with "x" 

```{r}
7*x
y <- 7*x

y

y/x
```

**Important shortcuts and advice**

- use Ctrl+R (Windows) or Command+Enter (Mac) to run a selection or a single line 
- everything after a pound sign (#) is a commentary 
- (i.e., R does not try to interpret numbers and character strings that follow a pound sign)

- always try to carefully comment your R-script/R-markdown files
- develop a clear style in writing R-script/R-markdown files
- guidelines for a clear style can be found at https://google.github.io/styleguide/Rguide.xml

- basic terminology for functions:     

```
        function(arg1, arg2, arg3, ...)    e.g.       sqrt(x = 36)
         |          |   |    |                         |   |   |
  function name     arguments            function name / argument name / argument value
                                            "sqrt"            "x"            "36"

```
### Help

```{r}
?sqrt # see help for sqrt()
```

## 2. Set Working Directory

- You need specify the corresponding path on your computer

```{r}
# find out current working directory
getwd()

# set working directory to my laptop desktop
setwd('/Users/nataliekupperman/Desktop') # working directory on my laptop

# set working directory back to One Drive
setwd("/Users/nataliekupperman/Library/CloudStorage/OneDrive-UniversityofVirginia/Courses/DS2003/DS-3003/01 Intro to Course & R/class 2") 
```


## 3. Reading Data

### Read TABULAR separated data

```{r}
# read in data
incex <- read.table('../data/income_exmpl.dat', header = T, sep = "\t") # use (..) to move up a directory

# print the first 6 lines of data with function head()
head(incex)       

# if the data are not located in the working directory (see setwd()) then you can read the data by specifying the full path
incex <- read.table('/Users/nataliekupperman/Library/CloudStorage/OneDrive-UniversityofVirginia/Courses/DS2003/DS-3003/01 Intro to Course & R/data/income_exmpl.dat', header = T, sep = "\t")

# if an R function requires a character string or name as an argument (e.g., the file name or path), it must be set in quotes ("name" or 'name')
```

### Read COMMA separated data

```{r}
# help on reading CSV data
?read.csv 

incex.2 <- read.csv('../data/income_exmpl.csv')
```

### Read other data formats (e.g., SPSS, SAS, or STATA)
 - use the 'foreign'-package 
 - install package from server, for example, in Berkeley 
 - alternatively use menu bar, `Tools`
 
```{r}
#install.packages('foreign', repos = 'http://cran.cnr.Berkeley.edu/') 

# load package (makes its functions available)
library(foreign) 

# help on the package
library(help = 'foreign')  

# help on reading SPSS data files (.sav)
?read.spss  

incex.3 <- read.spss('..data/income_exmpl.sav', to.data.frame = TRUE)

```

**Advice**

- If you want to understand an R function you need to carefully read the help file.
- This saves you from getting cryptic error messages and becoming increasingly frustrated.
- In particular, look at the default setting of the arguments, e.g. for to.data.frame

## 4. Generic Functions

### Simple description of data structure
```{r}
# prints the data on the screen = print(incex)
incex 

# returns the first six lines of the data frame
head(incex, 10)   

# returns the last six lines of the data frame
tail(incex) 

# returns the first ten lines by subscripting: incex[rows, cols]
incex[1:10, ]    

# shows the structure of the data frame
str(incex)         

# dimension of data (rows and colums)
dim(incex)           

# number of rows
nrow(incex)   

# number of columns
ncol(incex)  

# variable names of the data frame
names(incex)     

# list of row & variable names of the data frame
dimnames(incex)      

 # type (class) of R-object: data.frame
class(incex)  

# extracts & prints the "age" variable from the data frame
incex$age  

# returns the first six lines of age variable
head(incex$age)     

# class of age variable: integer vector
class(incex$age)  

# length of the factor (number of observations)
length(incex$edu)    

# extracts & print the factor values on the screen
incex$edu          

# returns the first six lines of edu variable
head(incex$edu)    

# class of edu variable: factor (vector of categorical values)
class(incex$edu) 

# number of factor levels
nlevels(incex$edu)

# levels of the factor
levels(incex$edu)    

```

### Checking available objects & search path
```{r}
objects()            # lists the objects accessible to you; alternatively: ls()
search()             # lists attached packages, and R objects,

rm(incex.2, incex.3) # removes the two data sets
objects()
```

### Getting help

```
help(mean)           # get help on R-objects (mostly functions)
?mean                # alternative version

help.search('mean')  # if you don't know the name of the function (object)
??mean               # alternative version

apropos('mean')      # searches for objects (incl. functions) containing "mean" 
                     # in their name

????mean             # if you still cannot find what you want

help.start()         # help "menu"
```

### Creating a Vector of Sequences

```{r}
                     # creates a vector of numeric values
                     # combines (concatenates) single values to a vector
c(1, 2, 3)           # very frequently used (whenever you have to pass a vector to a function)!
          
c('apple', 'tree')   # combines character strings into a character vector
c(3, 'apple')        # transforms 3 into a character string ("3")

```

### - EXAMPLE 1: assign/store a vector of values
```{r}
x <- c(21, -3, .003)
x
(y <- c('X1', 'X2', 'Y1', 'Y2'))   
# putting the assingment in parentheses "()" is equivalent to print() ... print on screen 
```

### - EXAMPLE 2: extract variable 'age' and 'occ' from the incex data frame
```{r}
# incex[, c('age', 'occ')]  # character strings 'age' and 'occ' need to be combined
head(incex[, c('age', 'occ')])
```

### - EXAMPLE 3: extract rows and columns from a data frame 
  - use the sequence operator ':'
  
```{r}
1:3                  # sequence of integers from 1 to 3; is the same as c(1, 2, 3)
3:1                  # sequence of integers from 3 to 1; is the same as c(3, 2, 1)
-25:-78
?':'                 # get help for ':'; need quotes since the colon is a special symbol

incex[1:10, c('age', 'occ')]  
incex[1:10, c(2, 4)]
incex[c(1:10, 20:25), c(2, 4)]

c(1:10, 20:25)
```

### the "sequence" function seq()
```{r}
?seq
seq(from = 1, to = 3, by = 1)
seq(from = 1, by = 1, to = 3)
seq(1, 3, 1)         # don't need the argument names if we list arguments in default order

seq(-2, 3, by = .25)
seq(-2, 3, length = 20)
```

### - EXAMPLE 1: use seq() for generating quantiles
```{r}
?quantile
seq(0, 1, .25)
quantile(incex$income, probs = seq(0, 1, .25))   # quartiles
quantile(incex$income, probs = seq(0, 1, .1))    # deciles

seq(0, 1, .1)
```

### - EXAMPLE 2: use seq() for plotting histograms
```{r}
?hist
hist(incex$income, breaks = seq(500, 2500, 100)) # bin-width of 100 EUR 
hist(incex$income, breaks = seq(500, 2500, 50))  # bin-width of 50 EUR
```

### - EXAMPLE 3: use seq() for generating names (e.g., for factor levels or plotting)
```{r}
?paste
paste('age:', seq(15, 70, 5))        # paste() builds a vector of character strings 
paste('age: ', seq(15, 70, 5), '-', seq(19, 74, 5), ' years', sep = '')
                       # shorter vectors (e.g., 'age', '-', are automatically recycled!
```

### the "replicate" function rep()
```{r}
rep(7, times = 10)
rep(7, 10)
rep(1:3, 10)
rep(1:3, each = 10)
rep(1:3, c(2, 5, 7))
?rep

rep(c(5, -1, 3, 0), each = 3)
rep(c('one', 'two', 'three'), 1:3)   # works also with character strings
```

### - EXAMPLE: generate a combination of character strings (or create factor combinations)

```{r}
A <- paste(rep('a', 5), 1:5, sep = '')
```


```{r}
# two factors A and B with 5 and 10 factor levels each:
# a1, a2, ..., a5, and b1, b2, ..., b10
# first, create vector of factor levels:
(A <- paste(rep('a', 5), 1:5, sep = ''))
(A <- paste('a', 1:5, sep = ''))     # even shorter (R recycles 'a' five times)
(B <- paste('b', 1:10, sep = ''))

# second, create all possible combinations of A and B 


cbind(rep(A, each = 10), rep(B, 5))  # cbind() binds the two vectors into a column matrix
cbind(rep(A, each = 10), B)          # shorter: use the recycling rule

# create a vector of character strings (of combinations)
paste(rep(A, each = 10), rep(B, 5), sep = '*')
```


### Basic Data Manipulations

### Accessing a data frame (selecting columns & rows)

```
incex$age          # selects 'age'-variable using $
incex[, 'age']     # selects 'age'-variable using [..]
incex[, 2]         # selects 2nd column
```

```{r}
incex[1:10, ]        # selects rows 1 to 10
incex[1:10, 2]       # selects rows 1 to 10 of age
incex[1:10, 'age']   # selects rows 1 to 10 of age
incex$age[1:10]      # first select age then rows 1 to 10
```

### - EXAMPLE: select a random sample of 20 observations
```{r}
smpl.ind <- sample(1:nrow(incex), size = 20)
smpl.ind
incex[smpl.ind, ] 
# alternatively (without creating the smpl.ind object):
incex[sample(1:nrow(incex), size = 20), ]  
```

### Selecting subsets of data using indices
```{r}
age <- incex$age     # creates new object of name "age" (ie, creates a copy)
dim(age)             # does not work since age is no longer a data frame but a vector
length(age)
age[10:20]           # only one argument in brackets since a vector has only 
                     # one dimension (rows)
age[seq(1, length(age), by = 5)]  # select each 5th value (starting with first obs.)
age[-(10:1700)]      # all values except values #10 to #1700
age[-c(1:9, 21:length(age))]      # all values except ...
```

### Sorting data 
```{r}
                     # sort age from the smallest to largest value
                     # note: the age variable 'age' remains unsorted since we
sort(age)            # didn't save the sorted vector: age <- sort(age)
order(age)           # returns the sorting index (for sorting)
                     # the first element implies that the samllest one is located on...
rank(age)            # return the rank (with ties)
```

### Selecting subsets of a vector using logical vectors
```{r}
inc <- incex$income
age <- incex$age
age.ord <- order(age)
inc[age.ord[1:10]]               # income of the 10 youngest employees
inc[age == 18]                   # income of 18 years old employees
head(age == 18)                  # creates a logical vector (TRUE, FALSE) and return the first six elements
inc[age >= 60]                   # income of 60 year old or older employees
inc[age <= 30 | age >= 60]       # income of 30 year old or younger, or 60 years or older employees
```

### Selecting a subset of a data frame
```{r}
incex[age == 18, c('income', 'edu')]  # select income and educ. level for empl. of age 18

incex18 <- subset(incex, age == 18)   # creates a subset of incex 
                                      # (only of 18y old employees)
incex18

# also select variables
subset(incex, age == 18, c(age, edu, occ))
subset(incex, age == 18, age:occ)
```

### Assigning values to selected cells 
```{r}
age[1:10]
age[c(3, 7)] <- 0      # assign age of 0 to 3rd and 7th observation
age[1:10]
age[age < 18] <- NA    # assign a missing value (Not Available) to age values less than 18
age[1:10]
```


## 7. Logical Operators

```{r}
TRUE                 # TRUE
FALSE                # FALSE
!TRUE                # ! ... logical negation (NOT TRUE)
!FALSE               # ! ... logical negation (NOT FALSE)
1 == 1               # == ... EQUAL
1 == 2
1 != 2               # != ... NOT EQUAL
'apple' == 'apple'   # logical operations also work for character strings
'apple' == 'orange'
'apple' != 'orange'

c(3, 7, 8) == 7      # logical operation with vectors (elementwise)
7 == c(3, 7, 8)
c(3, 7, 8) == c(1, 3, 8)
c(3, 7, 8) == c(1, 3, 8, 5, 7, 3)  # automatically recycles shorter vector (no warning!!!)
c(3, 7, 8) == c(1, 3, 8, 5, 7)     # automatically recycles shorter vector (warning message)

TRUE & TRUE          # & ... logical AND
TRUE & FALSE
FALSE & FALSE
TRUE | TRUE          # | ... logical OR
TRUE | FALSE
FALSE | FALSE
1 == 1 & 'apple' == 'orange'
1 == 1 | 'apple' == 'orange'
```


## 8. Missing data

```{r}
# ::::: checking for missing values (NA: not available) & selecting complete data :::::
# is.na(incex)         # check for missing values "is not available (na)?"
head(is.na(incex))
any(is.na(incex))    # check whether there are any missing values
sum(is.na(incex))    # how many missing values? (TRUE=1 and FALSE=0)

# EXAMPLE: data frame with NAs
x.df <- data.frame(a = c(5, 3, 9, NA, 0), b = c(NA, 8, 7, 1, 1))  # generate a data frame
x.df
is.na(x.df)
any(is.na(x.df))
sum(is.na(x.df))
x.df.complete <- na.omit(x.df)  # deletes all rows with a missing (NA)
x.df.complete
```

## 9. Other Data Manipulations

### Change order of levels for edu & occ
```{r}
summary(incex)       # summary stats for each variable of a data frame 
incex$occ <- factor(incex$occ, levels = c('low', 'med.', 'high'))
incex$edu <- factor(incex$edu, levels = c('low', 'med.', 'high'))
incex$sex <- factor(incex$sex, levels = c('m', 'f'), labels = c('male', 'female'))
summary(incex)

head(incex$occ)
head(incex$edu)
incex$edu >= 'med.'  # does not work since no order is defined!

# create ordered factor levels (for ordinal variables)
incex$occ <- factor(incex$occ, levels = c('low', 'med.', 'high'), ordered = T)
incex$edu <- factor(incex$edu, levels = c('low', 'med.', 'high'), ordered = T)
head(incex$occ)
head(incex$edu)
incex$edu >= 'med.'
```

### Use the transform() or within() functions
```{r}
# read data again (in order to get original occ, edu and sex variables)
incex <- read.table('income_exmpl.dat', header = T, sep = "\t")
summary(incex)

incex <- transform(incex, 
   occ = factor(occ, levels = c('low', 'med.', 'high')),
   edu = factor(edu, levels = c('low', 'med.', 'high')),
   sex = factor(sex, levels = c('m', 'f'), labels = c('male', 'female'))
)
summary(incex)


# read data again (in order to get original occ, edu and sex variables)
incex <- read.table('income_exmpl.dat', header = T, sep = "\t")
summary(incex)

incex <- within(incex, {
   occ <- factor(occ, levels = c('low', 'med.', 'high'))
   edu <- factor(edu, levels = c('low', 'med.', 'high'))
   sex <- factor(sex, levels = c('m', 'f'), labels = c('male', 'female'))
})
summary(incex)
```

### Clean up - delete objects

```{r}
objects()
rm(x, y, age, inc, age.ord, incex18, x.df, x.df.complete)
objects()
```


## 10. Basic Computing with Data

### Simple statistics for a data vector
```{r}
table(incex$edu)    
mean(incex$age)      
median(incex$age)
range(incex$age)
quantile(incex$age)
summary(incex$age)
summary(incex$edu)

mean(incex$income)
mean(incex$income[incex$edu == 'low'])  # mean for a subset
tapply(incex$income, incex$edu, mean)   # means by educational level
tapply(incex$income, list(incex$edu, incex$occ), mean)  # by education & occ. status
tapply(incex$income, list(incex$edu, incex$occ), sd)    # corresp. standard dev.
```

### Use with() or attach() and detach()
```{r}
# alternatively use with() instead of repeated 'incex$"
with(incex, tapply(income, edu, mean))
with(incex, tapply(income, list(edu, occ), mean))

# or attach() and detach()
# income               # not found since "income" only exist within "incex"
attach(incex)        # makes all variable of the incex data frame available
head(income)
tapply(income, edu, mean)
tapply(income, list(edu, occ), mean)
detach(incex)        # after you are done, detach data set
# income

# !!! WARNING: in order to avoid confusions when manipulating data
# avoid using attach() and detach() !!! But attaching data is useful for plotting.     
```


## 11. Writing Simple Functions

```{r}
# ::::: mean value & standard dev. :::::
mean.val <- function(x)
{
   # computes mean value of vector x
   sum(x) / length(x)
}

mean.val             # without (), prints the body of the function

mean.val(incex$income)
mean(incex$income)

sd.val <- function(x)
{
   # computes standard deviation of vector x
   sqrt(sum((x - mean.val(x))^2) / (length(x) - 1))
}

sd.val(incex$income)
sd(incex$income)

# define your own my.summary() function
my.summary <- function(x)
{
   # returns mean, sd, n, and standard error
   # x ... numeric vector (no missings allowed)

   m <- mean.val(x)
   s <- sd.val(x)
   n <- length(x)
   se <- s / sqrt(n)  # standard error
   c(mean = m, std.dev = s, n = n, s.e. = se)
}

my.summary(incex$income)
```


## 12. Visualizing Data

### Histogram & kernel density estimation
```{r}
attach(incex)        # attaches a data frame such that variables (colums)
                     # are directly accessible by their name
                     # after you no longer need the attached data frame detach() it

hist(income)
hist(income, prob = T)
lines(density(income), col = 'red')  # kernel density estimate; lines() adds lines to the active plot

detach(incex)        # detaches the incex data frame


library(lattice)     # load "lattice"-package: more fancy histograms and density plots
                     # most convenient with formula method
histogram(~ income, data = incex)
histogram(~ income | sex, data = incex)        # plot income by sex
histogram(~ income | sex * edu, data = incex)  # plot income by sex x edu
densityplot(~ income, data = incex)
densityplot(~ income | sex, data = incex)
densityplot(~ income | sex * edu, data = incex)
```

### Barplot
```{r}
barplot(table(incex$occ), main = 'occupational status')
barplot(table(incex$edu), main = 'education')
plot(~ occ, data = incex)      # alternative way

table(incex$age)
plot(occ ~ age, data = incex)   # plot calls spineplot
spineplot(occ ~ age, data = incex)
cdplot(occ ~ age, data = incex) # conditional density plot

barplot(table(incex$occ, incex$edu), main = 'occupational status', xlab="edu", ylab="occ")
barplot(prop.table(table(incex$occ, incex$edu), 2), main = 'occupational status', xlab="edu", ylab="occ")
plot(occ ~ edu, data = incex)
plot(occ ~ edu + sex, data = incex)  # need mouse or keyboard entry to confirm page change

mosaicplot(occ ~ edu, data = incex)
mosaicplot(edu ~ occ, data = incex)
mosaicplot(occ ~ edu + sex, data = incex)
mosaicplot(occ ~ edu + sex, data = incex, color = c('blue', 'red'))
mosaicplot(~ occ + edu + sex, data = incex)  # alternative formula
```

### Scatter Plot
```{r}
plot(income ~ oexp, data = incex)
coplot(income ~ oexp | edu, data = incex)
coplot(income ~ oexp | edu * occ, data = incex)

# from the "lattice"-package
xyplot(income ~ oexp, data = incex)
xyplot(income ~ oexp | edu, data = incex)
xyplot(income ~ oexp | edu * occ, data = incex)
```

## 13. Basic Tables

```{r}
sapply(incex, mean)
sapply(incex, class)
sapply(incex, function(x) sum(is.na(x)))
apply(incex, 2, function(x) sum(is.na(x)))
apply(incex, 1, function(x) sum(is.na(x)))
```

### Create tables
```{r}
xtabs(~ occ + edu, data = incex)  # with formula method
xtabs(~ sex, data = incex)
xtabs(income ~ sex, data = incex) # sum of income
xtabs(income ~ sex, data = incex) / xtabs( ~ sex, data = incex) # mean income by sex
with(incex, tapply(income, sex, mean))
with(incex, tapply(income, list(occ, edu), mean))
```

### Compare group means
```{r}
with(incex, tapply(income, sex, summary))
with(incex, tapply(income, sex, mean))
with(incex, tapply(income, sex, sd))
with(incex, tapply(income, sex, sd) / sqrt(c(table(sex))))  # s.e.

boxplot(income ~ sex, data = incex, varwidth = T, horizontal = T)
boxplot(income ~ sex, data = incex, varwidth = F, horizontal = T)

t.test(income ~ sex, data = incex)
t.test(income ~ sex, data = incex, var.equal = T)
```

### Crosstables
```{r}
with(incex, table(sex))
tab <- with(incex, table(sex, edu))
tab
plot(sex ~ edu, data = incex)
chisq.test(tab)

tab <- with(incex, table(occ, edu))
tab
plot(occ ~ edu, data = incex)
chisq.test(tab)

tab <- with(incex, table(occ, edu, sex))
tab
addmargins(tab)         # add totals to table
ftable(tab)             # print a better formated table
ftable(addmargins(tab))
mosaicplot(occ ~ edu + sex, data = incex)
```


## 14. Basic Data Structures

### Vectors (numeric & character)
```{r}
(v <- c(3, -2, 13, 10.5))
is.vector(v)
is.matrix(v)
length(v)
v[3:4]

(v2 <- c('a', 'b', 'c', 'b'))
is.vector(v2)
length(v2)
v2[3:4]

names(v) <- names(v2) <- 1:4
v
v2
```

### Factor 
```{r}
(f <- factor(c(0, 1, 1, 0)))
(f <- factor(c(0, 1, 1, 0), labels = c('male', 'female')))
is.factor(f)
length(f)
f[3:4]

(f2 <- as.factor(v2))      # transform a character vector into a factor
```

### Matrix
```{r}
(m <- matrix(1:20, nrow = 4, ncol = 5))
is.matrix(m)
dim(m)
m[2:3, 3:4]

dimnames(m) <- list(1:4, paste('X', 1:5, sep = ''))
m
m[2:3, c('X3', 'X4')]

(m2 <- cbind(v, m))        # create a matrix by binding columns together
is.matrix(m2)
is.data.frame(m2)

(m3 <- cbind(v, v2, f, m)) # matrix of character strings
```

### Data frame
```{r}
(d <- data.frame(v, v2, f, m))
is.data.frame(d)
is.matrix(d)
dim(d)
d[2:3, c('X3', 'X4')]
d[, 'X3']                  # select variable X3
d$X3                       # alternative way to select X3
# m3$X3                      # does not work for matrices, only works for data frames

dimnames(d)                # names can be changed as for the matrix above
```

### List
```{r}
(l <- list(v, v2, f, m))   # list with four list elements
is.list(l)
length(l)
l[4]                       # fourth list element returned as a list of length 1 (containing a matrix)
l[[4]]                     # fourth list element returned as matrix

(l <- list(V1 = v, V2 = v2, sex = f, data = m))   # list with assigned names
l$sex
```
