---
title: "Assignment 13 - Shiny and Maps (4 pts)"
author: "Group #, NAME"
date: "Due Date: 11:59pm, Apr 4"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
# runtime: shiny    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(shiny)
library(shinythemes)
```

# Group Homework 

- You will work with your group to complete this assignment.

- Upload your Shiny app on [ShinyApps.io](https://www.shinyapps.io).

- Submit your group's shared `.Rmd` AND your URL (https://XXXXX.shinyapps.io/XXXXX) on Collab.

- Confirm this with the following comment included in your submission text box: **"Honor Pledge: I have recreated my group submission using using the tools I have installed on my own computer"**
    
- Name the files with a group name and YOUR name for your submission.

- One of the group members will **present R codes and plots for Parts 1 and 2** in class on **Apr. 4 (Mon)**. Please e-mail the instructor if you're a presenter by 11:59pm, Apr. 3. Also, if you're a presenter, please bring your laptop so that you can share your screen on zoom for the presentation.

# Part 1

- Get the latest data from from [https://covid19.who.int/info?openIndex=2](https://covid19.who.int/info?openIndex=2).
      
   - The file should likely be named "WHO-COVID-19-global-table-data.csv"
      
   - Don't use the data for the previous group assignments. It is not the most recent data. 

- Choose three WHO regions of interest. Note that we have 6 main regions: Africa, Americas, Eastern Mediterranean, Europe, South-East Asia, and Western Pacific.

- Create a subset including 3 countries per `WHO region`. You can choose any three countries within each `WHO region` to compare the mortality rate (`mutate(rate = "Deaths - cumulative total"/"Cases - cumulative total")`). You will have 9 countries (3 countries * 3 WHO regions).

- Using `navbarPage()`, create a shiny dashboard that contains 3 tabs where each tab has a barplot of mortality rates with error bars from three countries in the selected region. For example,

```
ui <- navbarPage(title = "Mortality Rate", 
      tabPanel("Africa", ...), 
      tabPanel("Americas", "...), 
      tabPanel("Eastern Mediterranean", ...))
```      
- Create separate `tabPanel()` for each level of`WHO region` variable.

- Under each tab, create a bar plot (with error bars) from three countries in the selected region with a controller for including/excluding errors bars. The default plot has no error bars.

- Use different control types under each tab. Check available control types from [Widget gallery](https://shiny.rstudio.com/gallery/widget-gallery.html).
  
- Change a theme from the `shinythemes` package. See options from https://rstudio.github.io/shinythemes/.   

- Tips for  `renderPlot()`:

```
    output$<id> = renderPlot({
     
     p1 <-  ...your ggplot...
     print(p1)
      
     if (input$<id>) { # modify this line
     p1_e <- p1 + geom_errorbar(aes(ymin=rate-1.96*SE, ymax=rate+1.96*SE), width=.2)  # add error bars
     print(p1_e)
     }}) 
```

- Example: https://youmi.shinyapps.io/A13part1/

```{r}
# add your codes 
```

# Part 2

- Create a choropleth US map like the Trump Impeachment example. You need to add any state-related variable (e.g., the number of universities per state, the crime rate per state) onto the US map.

- Use a dataset that is available in data repositories (e.g., [kaggle](https://www.kaggle.com/datasets)). But don't use the built-in datasets or datasets from packages. You need to find a state-related variable online. 

- Briefly describe the dataset you're using for this assignment (e.g., means to access data, context, sample, variables, etc...)
  
   + *describe your data.*
   
- Create a choropleth map that displays data on a map by shading states with different colors.

- You can use any map-related functions we've learned in class or any other available functions for creating maps. Your map can be a static map or an interactive map.

```{r}
# add your codes 
```

- Describe your choropleth map.

