---
title: "A14: Shiny with Maps"
author: "TYPE YOUR NAME HERE"
date: "Due Date: 11:59pm, Apr 3"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggmap)
```

## Individual Homework

- Upload your Shiny app on [ShinyApps.io](https://www.shinyapps.io). Submit your `.Rmd` AND your URL (https://XXXXX.shinyapps.io/XXXXX) on Collab.

- You need to create your favorite city map with a controller for zooming and the sidebar layout.

- If you like, you can change the output type from `ioslides_presentation` to `slidy_presentation`.  

- In your submission file, you should include either of the following statements:

   - "I pledge that I have neither given nor received help on this assignment."
   - "I have [received/given] help on this assignment [to/from] [student]; the nature of this support was [elaborate]."


## Individual Homework (Cont'd)

- Detailed instructions ([Example](https://youmi.shinyapps.io/A14example)):

    - Select your favoriate city, and use `get_map()` and `ggmap()` to draw your favoriate city map from Google Maps.

    - Create a shiny app with a slider (`sliderInput()`) that allows for changing the value for zooming. The min value is 5, the max value is 18, and the default value is 10.  
    
    - Use `sidebarLayout()` with `sidebarPanel()` and `mainPanel()` for your shinyapp layout.
    
    - Use `shinyApp(ui = ..., server =...)` to create your a shiny app. 

## Individual Homework (Cont'd)

- Don't include your API key in your `Rmd` file. Please delete your API key in your `Rmd` submission file after uploading your shiny app on [ShinyApps.io](https://www.shinyapps.io).

## Map Plot

```{r}
ui <- # add your codes

server <- # add your codes

shinyApp(ui = ui, server = server, options = list(height = 600)) # if you like, change the height value.
```
