---
title: "DS3003: Local Polynominal Regression & Loess"
subtitle: "R codes"
author: "Youmi Suk"
date: "2. 2. 2022"
output:
     html_document:
          toc: TRUE
          toc_depth: 4
          toc_float: true
          toc_collapsed: true
          theme: journal
          code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
knitr::opts_knit$set(global.par=TRUE, html.table.attributes = '')

library(ggplot2)
library(dplyr)
library(knitr)

# ::::: read income data :::::
incex <- read.table('/Users/youmisuk/Desktop/DS3003/W1 - Syllabus & R intro/income_exmpl.dat', header = T, sep = "\t")

# change order of factor levels
incex$occ <- factor(incex$occ, levels = c('low', 'med.', 'high'))
incex$edu <- factor(incex$edu, levels = c('low', 'med.', 'high'))
incex$sex <- factor(incex$sex, levels = c('m', 'f'), labels = c('male', 'female'))
```

## Local polynomial regression

- Write our own local linear or quadratic regression function: `k.reg()`
```{r}
k.reg <- function(y, x, h, deg = 1)
{
   # computes a predicted value of y for each unique value of x 
   # with a centered normal kernel
   # y   ... vector of dependent variable
   # x   ... vector of independent variable
   # h   ... bandwidth (standard deviation of the normal kernel
   # deg ... degree of polynomial (1: linear, 2: quadratic)

   x.val <- sort(unique(x))
   avrg <- function(y, x, x.loc, h, deg) {
      x.c <- x - x.loc        # centering
      wt <- dnorm((x.c) / h)  # weights
      if(deg == 1) m <- lm(y ~ x.c, weights = wt)$coef[1] # extracts intercept
      if(deg == 2) m <- lm(y ~ x.c + I(x.c^2), weights = wt)$coef[1]
      m                       # return mean value
   }
   m.vec <- sapply(x.val, avrg, y = y, x = x, h = h, deg = deg) 
   data.frame(x = x.val, y = m.vec)
}
```

- add a local reg. line onto a scatterplot.

```{r}
# bandwidth = 1
out.kreg <- with(incex, k.reg(income, oexp, 1))            # get x and local mean values
plot(income ~ oexp, data = incex, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', main = 'Local Polynomial Regression')
lines(out.kreg, col = 'red', lwd = 3)                      # add path of means
```

- add multiple reg. lines onto a scatterplot.

```{r}
plot(income ~ oexp, data = incex, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', main = 'Local Polynomial Regression')
lines(out.kreg, col = 'red', lwd = 3)                      # add path of means

# add paths of means with different bandwidth or degree of polynomials
out.kreg2 <- with(incex, k.reg(income, oexp, 3))           # bandwidth of 3
lines(out.kreg2, col = 'blue', lwd = 3)
out.kreg3 <- with(incex, k.reg(income, oexp, 3, deg = 2))  # quadratic polynomial
lines(out.kreg3, col = 'green', lwd = 3)

legend('topleft', c('Band = 1', 'Band = 3', 'Band = 3 + Quad'),
       col = c('red', 'blue', 'green'), lwd=rep(3, 3), inset = .05)
```

- If you like to use package `ggplot2`,

```{r}
colors <- c("Band = 1 & Linear" = "red", "Band = 3 & Linear" = "blue", "Band = 3 & Qaud" = "green")


ggplot(incex, aes(x=oexp, y=income)) + geom_point() + 
  labs(x = 'Occupational Experience', y = 'Monthly Net Income', title='Local Polynomial Regression', color = "Bandwidth & Degree") +    
  geom_line(data=out.kreg, aes(x=x, y=y, col="Band = 1 & Linear"), size=1) + 
  geom_line(data=out.kreg2, aes(x=x, y=y, col="Band = 3 & Linear"), size=1) + 
  geom_line(data=out.kreg3, aes(x=x, y=y, col="Band = 3 & Qaud"), size=1) +
  scale_color_manual(values = colors) + theme_bw()
```


## `loess()` function

- uses a span instead of a fixed bandwidth, i.e., the bandwidth at each 
- distinct x-value depends on the number of observations in the neighborhood (bandwidth is computed such that 20% of data are covered)

### - Local regression with varying degrees

```{r}
plot(income ~ oexp, data = incex, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', main = 'loess (LOcal regrESSion)')         # scatterplot

x.val <- seq(0, 48, by = .5)

# ::::: local regression (loess): but with simple means since degree = 0, ie kernel estimation :::::
out.lss <- loess(income ~ oexp, data = incex, span = .2, degree = 0)         # estimate model
y.pred <- predict(out.lss, data.frame(oexp = x.val))  # get predicted values
lines(x.val, y.pred, col = 'red', lwd = 2)

# ::::: local LINEAR regression (loess): degree = 1 :::::
out.lss1 <- loess(income ~ oexp, data = incex, span = .2, degree = 1)
y.pred <- predict(out.lss1, data.frame(oexp = x.val))
lines(x.val, y.pred, col = 'blue', lwd = 2)

# ::::: local QUADRATIC regression (loess): degree = 2 :::::
out.lss2 <- loess(income ~ oexp, data = incex, span = .2, degree = 2)
y.pred <- predict(out.lss2, data.frame(oexp = x.val))
lines(x.val, y.pred, col = 'green', lwd = 2)

legend('topleft', c('Span = .2 & Deg. = 0', 'Span = .2 & Deg. = 1', 'Span = .2 & Deg. = 2'),
       col = c('red', 'blue', 'green'), lwd=rep(2, 3), inset = .05)
```



### - Separate nonparametric regressions (for male and female employees)

```{r}
plot(income ~ oexp, data = incex, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', bty = 'L')                  # scatterplot
out.all <- loess(income ~ oexp, incex, span = .5, degree = 2) # estimate model for all employees
out.m <- loess(income ~ oexp, incex, subset = sex == 'male', 
         span = .5, degree = 2)                               # estimate model for men
out.f <- loess(income ~ oexp, incex, subset = sex == 'female', 
         span = .5, degree = 2)                               # estimate model for women
x.val <- seq(0, 48, by = .5)                                  # get x values where to estimate means
lines(x.val, predict(out.all, x.val), lwd = 3, lty = 2)       # predict & plot mean values for all
lines(x.val, predict(out.m, x.val), col = 'blue', lwd = 2)    # predict & mean values for men
lines(x.val, predict(out.f, x.val), col = 'red', lwd = 2)     # predict & mean values for women
```

## function `geom_smooth()` in `ggplot2`

- default span = 0.75 

```{r}
ggplot(incex, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income') +
        geom_smooth(method='loess', formula= y~x, col="red", se = FALSE) + theme_bw()
```

- try different spans.

```{r}
ggplot(incex, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income') +
        geom_smooth(method='loess', formula= y~x, span=0.75, col="red", se = FALSE) +
        geom_smooth(method='loess', formula= y~x, span=0.4, col="blue", se = FALSE) + 
        geom_smooth(method='loess', formula= y~x, span=0.2, col="green", se = FALSE) + theme_bw()
```

- with a color separator

```{r}
ggplot(incex, aes(x=oexp, y=income, col=sex)) + geom_point(alpha=0.4) + labs(x = 'Occupational Experience', y = 'Monthly Net Income') +
        geom_smooth(method='loess', formula= y~x, se = FALSE) + theme_bw()
```


## Comparison with linear regression 

```{r}
# ::::: comparison with linear regression :::::
plot(income ~ oexp, data = incex, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', bty = 'L')             # scatterplot
lines(x.val, y.pred, col = 'blue', lwd = 2)              # add loess() estimates
abline(lm(income ~ oexp, incex), lwd = 3, col = 'red')   # add linear regression line
```

