---
title: "DS3003: Kernel Estimation"
subtitle: "R codes"
author: "Youmi Suk"
date: "2. 2. 2022"
output:
     html_document:
          toc: TRUE
          toc_depth: 4
          toc_float: true
          toc_collapsed: true
          theme: journal
          code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=4)
knitr::opts_knit$set(global.par=TRUE, html.table.attributes = '')

library(ggplot2)
library(dplyr)
library(knitr)

# ::::: read income data :::::
dat <- read.table('/Users/youmisuk/Desktop/DS3003/W1 - Syllabus & R intro/income_exmpl.dat', header = T, sep = "\t")
summary(dat)

# change order of factor levels
dat$occ <- factor(dat$occ, levels = c('low', 'med.', 'high'))
dat$edu <- factor(dat$edu, levels = c('low', 'med.', 'high'))
dat$sex <- factor(dat$sex, levels = c('m', 'f'), labels = c('male', 'female'))

set.seed(1)
# create x and y variables.
n <- 200   # number of observations
x <- c(rnorm(n/2, -25, 10), rnorm(n/2, 25, 10))   # generate bimodally distributed variable
y <- 150 + 1.5*x + .01*x^2 - .001*x^3 + rnorm(n, mean = 0, sd = 10) 
```

## Scatterplots

* check the data

```{r}
# check the first 6 lines
head(dat)

# summary 
summary(dat)
```


* use default functions to draw a scatterplot.

```{r}
# scatterplot
plot(income ~ oexp, data = dat, cex = .6, xlab = 'Occupational Experience', ylab = 'Monthly Net Income')
```

* use ggplot to draw a scatterplot, regression smoother, and linear regression line.

```{r}
ggplot(dat, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income') + theme_bw()
```

## Conditional distribution & expectation

* use default functions.

```{r}
# conditional distribution
plot(income ~ factor(oexp), data = dat, cex = .4, xlab = 'Occupational Experience', 
     ylab = 'Monthly Net Income', main = 'Conditional Distribution')

# conditional expectation
plot(income ~ oexp, data=dat, cex = .4, xlab = 'Occupational Experience', 
     ylab = 'Monthly Net Income', main = 'Conditional Mean')

m.vec <- with(dat, tapply(income, oexp, mean))   # creates vector of conditional means
o.vec <- with(dat, sort(unique(oexp)))           # determines the unique values of oexp and sorts them

points(o.vec, m.vec, col = 'red', pch = 16, type = 'o', lwd = 2) # plot path of means
```

```{r}
# compute conditional quartiles
q50.vec <- with(dat, tapply(income, oexp, median))
q25.vec <- with(dat, tapply(income, oexp, quantile, prob = .25))  # first quartile
q75.vec <- with(dat, tapply(income, oexp, quantile, prob = .75))  # third quartile

plot(income ~ oexp, data=dat, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', main = 'Conditional Means/Distribution')
points(o.vec, m.vec, col = 'red', pch = 16, type = 'l', lwd = 2) # plot path of means
points(o.vec, q50.vec, col = 'blue', pch = 16, type = 'l', lwd = 2)
points(o.vec, q25.vec, col = 'blue', pch = 16, type = 'l', lwd = 2, lty = 2)
points(o.vec, q75.vec, col = 'blue', pch = 16, type = 'l', lwd = 2, lty = 2)
legend('topleft', c('mean', 'median', '1st quartile', '3rd quartile'),
       col = c('red', rep('blue', 3)), lwd = 2, lty = c(1, 1, 2, 2), inset = .05)

```

* use ggplot.

```{r}
# conditional distribution
ggplot(dat, aes(x=oexp, y=income)) + geom_boxplot(aes(group=oexp)) + labs(x = 'Occupational Experience', y = 'Monthly Net Income', title='Conditional Distribution') + theme_bw()

# conditional expectation
ggplot(dat, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income', title='Conditional mean') + theme_bw() + 
   stat_summary(fun = mean, geom = 'point', col = "red", size=2.5) + 
   stat_summary(fun = mean, geom = 'line', col = "red", size=1) 
```

## Binning

* use default functions 

```{r}
# create equi-distant bins using cut() (results in a "factor")
dat$oexpbins <- cut(dat$oexp, breaks = seq(0, 50, by = 5), right = F) 
m.vec <- with(dat, tapply(income, oexpbins, mean))      # vector of income means for each bin
o.vec <- with(dat, tapply(oexp, oexpbins, mean))        # vector of oexp means for each bin

# plot scatterplot and path of bin-means
plot(income ~ oexp, data = dat, cex = .4, xlab = 'Occupational Experience', 
     ylab = 'Monthly Net Income', main = 'Path of Group Means')
points(o.vec, m.vec, type = 'o', col = 'red', lwd = 2)  # plot path of means
abline(v = seq(4.5, 50, by = 5), lty = 2)               # add bin borders

# plot scatterplot and stepfunction of bin-means
plot(income ~ oexp, data = dat, cex = .4, xlab = 'Occupational Experience', 
     ylab = 'Monthly Net Income', main = 'Step Function of Group Means')
points(seq(-.5, 50, by = 5), c(m.vec, m.vec[length(m.vec)]), type = 's', 
       col = 'red', lwd = 3)                            # type = 's' plots a stepfunction
```

* use ggplot

```{r}
# create equi-distant bins using cut() (results in a "factor")
dat$oexpbins <- cut(dat$oexp, breaks = seq(0, 50, by = 5), right = F) 
bin.stat <- dat %>% group_by(oexpbins) %>%  
            summarize (m.income = mean(income), m.oexp = mean(oexp), .groups = "drop") # income means and oexp means for each bin

# plot scatterplot and path of bin-means
ggplot(dat, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income', title="Path of Group Means") + theme_bw() + 
   geom_point(data=bin.stat, aes(x=m.oexp, y=m.income), col = "red", size=2.5) + 
   geom_line(data=bin.stat, aes(x=m.oexp, y=m.income), col = "red", size=1) + 
   geom_vline(xintercept = seq(0, 50, by = 5), linetype="dotted", size=0.5)

# plot scatterplot and stepfunction of bin-means
bin.stat2 <- cbind(rbind(bin.stat, bin.stat[nrow(bin.stat),]), breaks=seq(0, 50, by = 5))
ggplot(dat, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income', title="Step Function of Group Means") + theme_bw() + 
   geom_step(data=bin.stat2, aes(x=breaks, y=m.income), col = "red", size=1)
```

## Local averaging

* write our own function for local averaging: loc.av()

```{r}
loc.av <- function(y, x, w)
{
   # computes local means of y for each unique value of x 
   # with a centered/symmetric window of width w
   # y ... vector of dependent variable
   # x ... vector of independent variable
   # w ... width of the window for averaging

   x.val <- sort(unique(x))          # determines the unique values of x and sorts them
   avrg <- function(y, x, x.loc, w) {
      # determines the local mean value at a specific value x.loc
      mean(y[x >= x.loc-w/2 & x <= x.loc+w/2])
   }
   m.vec <- sapply(x.val, avrg, y = y, x = x, w = w)  # apply function to each x.val
   data.frame(x = x.val, y = m.vec)                        # return data frame of x values and means
}
```

```{r}
out.av <- with(dat, loc.av(income, oexp, w = 5))            # matrix of x and mean values

plot(income ~ oexp, data = dat, cex = .4, xlab = 'Occupational Experience', 
     ylab = 'Monthly Net Income', main = 'Local Averaging')   # plot scatterplot
lines(out.av, col = 'red', lwd = 3)                           # add path of means obtained from loc.av()

# add path of means resulting from different window choices (10y and 3y)
out.av2 <- with(dat, loc.av(income, oexp, w = 10))
lines(out.av2, col = 'blue', lwd = 2)
out.av3 <- with(dat, loc.av(income, oexp, w = 3))
lines(out.av3, col = 'green', lwd = 2)

legend('topleft', c('window = 3', 'window = 5', 'window = 10'),
       col = c('green', 'red', 'blue'), lwd=c(2, 3, 2), inset = .05)
```

* use ggplot.

```{r}
out.av <- with(dat, loc.av(income, oexp, w = 5))     # matrix of x and mean values

ggplot(dat, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income', title='Local Averaging') + geom_line(data=out.av, aes(x=x, y=y), col = "red", size=1) + theme_bw()
```

## Kernel estimation

* write our own function for regression smoothing: k.smth()

```{r}
k.smth <- function(y, x, h)
{
   # computes a weighted mean of y for each unique value of x 
   # with a centered normal kernel
   # y ... vector of dependent variable
   # x ... vector of independent variable
   # h ... bandwidth (standard deviation of the normal kernel)

   x.val <- sort(unique(x))         # sorted unique values of x
   avrg <- function(y, x, x.loc, h) {
      # computes local mean value at x.loc using normal kernel weights
      wt <- dnorm((x - x.loc) / h)  # weigths
      (y %*% wt) / sum(wt)          # locally weighted mean; or: weighted.mean(y, wt)
   }
   m.vec <- sapply(x.val, avrg, y = y, x = x, h = h)   # applies avrg() to all x.vals
   data.frame(x = x.val, y = m.vec)      # returns matrix of x and mean values
}
```

```{r}
# bandwidth (standard dev. = 1)
out.smth <- with(dat, k.smth(income, oexp, 1))               # get x and mean values
plot(income ~ oexp, data = dat, cex = .4, xlab = 'Occup. Experience', 
     ylab = 'Monthly Net Income', main = 'Kernel Estimation')  # scatterplot
lines(out.smth, col = 'red', lwd = 3)                          # add path of means

# add path of means with a larger bandwidth (standard dev. = 3)
out.smth2 <- with(dat, k.smth(income, oexp, 3))
lines(out.smth2, col = 'blue', lwd = 2)

legend('topleft', c('SD = 1', 'SD = 3'),
       col = c('red', 'blue'), lwd=c(3, 2), inset = .05)
```

* use ggplot.

```{r}
out.smth <- with(dat, k.smth(income, oexp, 1))               # get x and mean values

ggplot(dat, aes(x=oexp, y=income)) + geom_point() + labs(x = 'Occupational Experience', y = 'Monthly Net Income', title='Kernel Estimation') + geom_line(data=out.smth, aes(x=x, y=y), col = "red", size=1) + theme_bw()
```

