---
title: "Visualizing Errors: Error Bars"
author: "Youmi Suk"
date: "3. 13. 2022"
output:
     html_document:
          toc: TRUE
          toc_depth: 4
          toc_float: true
          toc_collapsed: true
          theme: journal
          code_folding: show
---


```{r setup, include=FALSE}
knitr::opts_knit$set(global.par=TRUE, html.table.attributes = '')
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      fig.align = "center",
                      out.width = "80%")
options(width=10000)

library(dplyr)
library(knitr)
library(tidyverse)
library(DT)

path1 <- "/Users/youmisuk/Desktop/DS3003/W8 - Plotly & Plots with CI/"
path2 <- "/Users/youmisuk/Desktop/DS3003/Assign/"
```


## Bar plots with COVID data

### Bar plots

```{r}
covids <- read_csv(paste0(path1, "WHO-COVID-19-global-table-data.csv"))
DT::datatable(covids, options = list(scrollX = T))
```

```{r}
focus <- c('Sweden', 'United States of America', 'Denmark', 'Finland', 'Norway', 
           'Republic of Korea', 'Japan', 'China', 'Viet Nam', 'New Zealand')

covids %>% filter(Name %in% focus) %>% mutate(rate = `Deaths - cumulative total`/`Cases - cumulative total`) %>% 
ggplot(aes(x=rate, y=Name)) + geom_col() + theme_bw() + xlab('WHO: # Reported Deaths / # Reported Cases')
```


### Error bars

- Theoretical uncertainty in an estimate of a proportion

1. $\textrm{First we use} \; Variance \; \textrm{of a coin flip with probability p:}\; p(1-p)$

2. $\textrm{Then we calculate the} \; Variance \; \textrm{of an average of coin flips:}\; p(1-p)/n$

3. $\textrm{Then we calculate the} \; Standard \; Deviation \; \textrm{of an average of coin flips:}\; \sqrt{p(1-p)/n}$

4. $\textrm{And when you estimate this with the observed sample proprotion it's called the} \; Standard \; Error:\; \sqrt{\hat p(1-\hat p)/n}$

```{r}
covids2 <- covids %>% filter(Name %in% focus) %>% 
  mutate(rate = `Deaths - cumulative total`/`Cases - cumulative total`, 
         SE = sqrt(rate*(1-rate)/`Cases - cumulative total`))

hist(covids2$rate) # histogram of death rates
```

- Add error bars.

```{r}
ggplot(covids2, aes(x=rate, y=Name)) + geom_col() + theme_bw() + 
  xlab('WHO: # Reported Deaths / # Reported Cases') + 
  geom_errorbar(aes(xmin=rate-1.96*SE, xmax=rate+1.96*SE), width=.2)
```

- What is driving the degree of uncertainty?
  - Do we trust these estimates?
  - How could death rates should vary by country?
  
## Bar plots with incex data

- Load data

```{r}
# load incex data
incex <- read.table(paste0(path2, 'income_exmpl.dat'), header = T, sep = "\t")

# change order of factor levels
incex$occ <- factor(incex$occ, levels = c('low', 'med.', 'high'))
incex$edu <- factor(incex$edu, levels = c('low', 'med.', 'high'))
incex$sex <- factor(incex$sex, levels = c('m', 'f'), labels = c('male', 'female'))
```

- First, summarize the data. This can be done in many ways, as described on [R cookbook](http://www.cookbook-r.com/Manipulating_data/Summarizing_data/). In class, we’ll use the `summarySE()` function below.

```{r}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}
```

- Get standard errors and marginal errors. 

   - Margin of error: distance between the point estimate and the lower or upper confidence limit.

```{r}
incex_grp <- summarySE(incex, measurevar="income", groupvars=c("occ","sex"))
str(incex_grp)
kable(incex_grp)
```

- Use standard error bars of the mean

```{r}
ggplot(incex_grp, aes(x = occ, y=income, fill=sex)) + 
    geom_bar(position=position_dodge(), stat="identity") +
    geom_errorbar(aes(ymin=income-se, ymax=income+se),
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9))
```

-  Use marginal errors instead of standard errors to construct 95% confidence intervals.

```{r}
ggplot(incex_grp, aes(x = occ, y=income, fill=sex)) + 
    geom_bar(position=position_dodge(), stat="identity", col="black", size=.2) +
    geom_errorbar(aes(ymin=income-ci, ymax=income+ci),
                  width=.2, # Width of the error bars
                  size=.3, 
                  position=position_dodge(.9))
```

- The graph can be improved using additional functions (e.g., `labs`, `theme_...`, `scale_...`) as:

```{r}
ggplot(incex_grp, aes(x = occ, y=income, fill=sex)) + 
    geom_bar(position=position_dodge(), stat="identity",
             colour="black", # Use black outlines,
             size=.3) +      # Thinner lines
    geom_errorbar(aes(ymin=income-ci, ymax=income+ci),
                  size=.3,    # Thinner lines
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9)) +
    labs(x="Occupational status", y="Mean income", title="Gender pay gap by occupation status with 95% confidence intervals") +
    scale_fill_hue(name="Gender", # Legend label, use darker colors
                   breaks=c("male", "female"),
                   labels=c("Male", "Female")) +
    theme_bw()
```

## Line plots

- We could summarize the above data using a line graph. That is, we could use basic line and point graph with error bars representing either the standard error of the mean, or 95% confidence interval.

```{r}
# Standard error of the mean
ggplot(incex_grp, aes(x = occ, y=income, col=sex, group=sex)) + 
    geom_line() + geom_point() + 
    geom_errorbar(aes(ymin=income-se, ymax=income+se),
                  width=.1)
```

```{r}
# Standard error of the mean
ggplot(incex_grp, aes(x = occ, y=income, col=sex, group=sex)) + 
    geom_line() + geom_point() + 
    geom_errorbar(aes(ymin=income-ci, ymax=income+ci),
                  width=.1)
```

- Use a subset of 50 cases.

```{r}
set.seed(1)
incex2 <- sample_n(incex, 50, replace=FALSE) # randomly sample 50 cases.

incex_grp2 <- summarySE(incex2, measurevar="income", groupvars=c("occ","sex"))
str(incex_grp2)
kable(incex_grp2)
```

- Draw error bars.

```{r}
ggplot(incex_grp2, aes(x = occ, y=income, col=sex, group=sex)) + 
    geom_line() + geom_point() + 
    geom_errorbar(aes(ymin=income-ci, ymax=income+ci),
                  width=.1)
```

- To avoid overlapped errorbars, use `position_dodge` to move them horizontally.

```{r}
ggplot(incex_grp2, aes(x = occ, y=income, col=sex, group=sex)) + 
    geom_line(position=position_dodge(0.1)) + 
    geom_point(position=position_dodge(0.1)) + 
    geom_errorbar(aes(ymin=income-ci, ymax=income+ci),
                  width=.1,
                  position=position_dodge(0.1))
```

- Change the size of data points and change the color of error bars. 

```{r}
ggplot(incex_grp2, aes(x = occ, y=income, col=sex, group=sex)) + 
    geom_line(position=position_dodge(0.1)) + 
    geom_point(position=position_dodge(0.1), size=3) + 
    geom_errorbar(aes(ymin=income-ci, ymax=income+ci),
                  colour="black",
                  width=.1,
                  position=position_dodge(0.1))
```
