---
title: "DS3003: plotly"
author: "Youmi Suk"
date: "3. 13. 2022"
output:
     html_document:
          toc: TRUE
          toc_depth: 4
          toc_float: true
          toc_collapsed: true
          theme: journal
          code_folding: show
---

<!--- Change font size for headers --->
<style>
h1.title {
  font-size: 30px;
}
h1 {
  font-size: 26px;
}
h2 {
  font-size: 22px;
}
h3 { 
  font-size: 18px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA, message = FALSE)
knitr::opts_knit$set(global.par=TRUE, html.table.attributes = '')
library(tidyverse)

par(mar=c(2.1,2.1,.5,1.5), mgp=c(1.3,.3,0), tck=-0.01)

# ::::: read income data :::::
incex <- read.table('/Users/youmisuk/Desktop/DS3003/W1 - Syllabus & R intro/income_exmpl.dat', header = T, sep = "\t")
summary(incex)

# change order of factor levels
incex$occ <- factor(incex$occ, levels = c('low', 'med.', 'high'))
incex$edu <- factor(incex$edu, levels = c('low', 'med.', 'high'))
incex$sex <- factor(incex$sex, levels = c('m', 'f'), labels = c('male', 'female'))
```


## Plotly

- `plotly` is a high-level interface to [plotly.js](https://github.com/plotly/plotly.js), based on [d3.js](https://d3js.org/) which provides an easy-to-use UI to generate slick D3 interactive graphics. These interactive graphs give the user the ability to zoom the plot in and out, hover over a point to get additional information, filter to groups of points, and much more. These interactive components contribute to an engaging user experience and allows information to be displayed in ways that are not possible with static figures.

- `plotly` is a web application for creating and sharing data visualizations. Plotly can work with several programming languages and applications including R, Python, and Microsoft Excel. We're going to concentrate on creating different graphs with `Plotly`.

- Interactive Graphics!
  - Zooming
  - Silencing
  - Hovering
  - Sliding, etc.

- If you want to share your visualizations on https://plot.ly/ you should make an account on their site.


### plot_ly Resources

- [The Plolty Website](https://plotly.com/)
- [The Plotly R documentation](https://plotly.com/r/)
- [The Plotly R Package on GitHub](https://github.com/plotly/plotly.R)
- [The Plotly R Cheatsheet](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf)

### How to create a plotly object

- There are two main approaches to initialize a plotly object:  
  
   + transforming a ggplot2 object with `ggplotly()` 
   + setting up aesthetics mappings with `plot_ly()` directly

## Web-based ggplot2 graphics: ggplotly

- `ggplotly()` takes existing `ggplot2` objects and converts them into interactive `plotly` graphics. That is, `ggplotly()` converts your static plots to an interactive web-based version!

- This makes it easy to create interactive figures because we are already familiar to the `ggplot2` syntax. 

### Bar Plot

```{r}
library(plotly)
g <- ggplot(data=incex) + geom_bar(mapping = aes(x = occ))

ggplotly(g)
```

### Histogram

```{r}
g2 <- ggplot(incex) + geom_histogram(aes(x = income), bins = 20) + theme_minimal()

ggplotly(g2)
```


### Density Plot

```{r}
g3 <- ggplot(incex) + geom_density(aes(x = income)) + theme_minimal()

ggplotly(g3)
```

### Scatterplot

```{r}
g4 <- ggplot(incex, aes(x = age, y = income)) + geom_point() + geom_smooth(method="loess", formula=y~x, se=F) + theme_bw()

ggplotly(g4)
```

## Using plotly without ggplot2

- `plot_ly()` is the base `plotly` command to initialize a plot from a dataframe, similar to `ggplot()` from ggplot2.


### Bar plot

- Specify `type = "bar"` to create a bar plot.

- More info from https://plotly.com/r/bar-charts/; https://plotly.com/r/reference/#bar

```{r}
occ_counted <- incex %>% count(occ, name = 'count') 
occ_counted

plot_ly(occ_counted, x= ~occ, y=~count, type = "bar")
```

- Add hover texts to a bar plot, and produce red bars with black outline.

```{r}
occ_counted$text <- c("low level", "medium level", "high level")

p<- plot_ly(occ_counted, x= ~occ, y=~count, type = "bar", 
             text = ~text, 
              color = I("skyblue"), 
              stroke = I("red"), 
              span = I(10))
p
```

```{r}
# change title, xlabel, and ylabel
p %>% layout(title = "Freq. of Occupation Status",
         xaxis = list(title = "Occupation Status"),
         yaxis = list(title = "Frequency"))
```

### Boxplot

- Specify `type = "box"` to create a boxplot.

- More info from https://plotly.com/r/box-plots/

```{r}
plot_ly(incex, y = ~income, type = "box", name = '')
```

- create grouped boxplots.

```{r}
?airmiles
plot_ly(incex, y = ~income, color = ~occ, type = "box")
```

### Histogram

- Use the `type = "histogram"` argument.

- More info from https://plotly.com/r/histograms/

```{r}
plot_ly(incex, x = ~income, type = "histogram")
```

### Scatterplot

- Specify a scatterplot by indicating `type = "scatter"` and `mode = "markers"`.

- More info from https://plotly.com/r/line-and-scatter/; https://plotly.com/r/reference/#scatter; https://plotly.com/r/reference/#scatter-mode

```{r}
plot_ly(incex, x = ~income, y = ~age, type = "scatter", mode = "markers")
```

#### Scatterplot with different colors

- Add color to your scatterplot points according to a categorical variable in the data frame you use with `plot_ly()`.

```{r}
plot_ly(incex, x = ~income, y = ~age, type = "scatter", mode = "markers",
        color = ~edu)
```

- You may want to change a color brewer palette.

```{r}
plot_ly(incex, x = ~income, y = ~age, type = "scatter", mode = "markers",
        color = ~edu, colors = "Set1")
```

#### Scatterplot with different symbols

- Create a subset of incex data. Change symbols for educational status.

```{r}
set.seed(1)
index <- sample(1:nrow(incex), 100, replace=FALSE)

subincex <- incex[index, ]
plot_ly(subincex, x = ~income, y = ~age, type = "scatter", mode = "markers",
        symbol = ~edu, 
        symbols = c('circle','x','o'), 
        color = I('black'), 
        marker = list(size = 10))
```

### 3D Scatterplot

- Create a three-dimensional scatterplot with the `type = "scatter3d"` argument. If you click and drag these scatterplots, you can view them from different angles.

```{r}
plot_ly(subincex, x = ~income, y = ~age, z= ~oexp, type = "scatter3d", mode = "markers", color = ~oexp, name="")
```

### Lineplot

- Specify a scatterplot by indicating `type = "scatter"` and `mode = "lines"`.

- More info from https://plotly.com/r/line-charts/; https://plotly.com/r/reference/#scatter; https://plotly.com/r/reference/#scatter-mode 

- Line graphs are useful for showing change over time. 

- For demonstration, use `airmiles` data that is time series data. 


```{r}
data("airmiles") # load airmiles data
str(airmiles) # check structure

plot_ly(x = time(airmiles), y = airmiles, type = 'scatter', mode = 'lines')
```

