---
title: "Assignment 16 - Animations (4 pts)"
author: "Group #, NAME"
date: "Due Date: 11:59pm, Apr 10"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
# runtime: shiny    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(shiny)
library(shinythemes)
```

# Group Homework 

- You will work with your group to complete this assignment.
.
- Submit your group's shared `.Rmd` AND your kintted `html` on Collab.

- You **don't** need to upload the html file on RPubs. 

- Confirm this with the following comment included in your submission text box: **"Honor Pledge: I have recreated my group submission using using the tools I have installed on my own computer"**
    
- Name the files with a group name and YOUR name for your submission.


# Part 1: `gganimate`

- First, get the latest data from [OWID](https://ourworldindata.org/covid-deaths).
      
   - The file should likely be named "owid-covid-data.xlsx" or "owid-covid-data.csv." (or jason file)
      
- Second, choose six countries that are of interest to you, and create a subset with the six countries.

- Third, create an animated line plot of "Covid Deaths Each Day Per 100,000 People in Countries" as in our lecture note on Animations.  


```{r}
# add your codes 
```

- Fourth, describe your findings from the plot.


# Part 2: `plotly`

- Create an animated plot with the `Play` button from package `plotly`. Please refer to our lecture note on Animations, [feature of plotly](https://plotly.com/r/animations/), and [cumulative anmiations](https://plotly.com/r/cumulative-animations/).

- First, use an appropriate dataset that is available in data repositories (e.g., [kaggle](https://www.kaggle.com/datasets)). But don't use the built-in datasets or datasets from packages. Don't use any datasets you used for the previous assignments.

- Second, briefly describe the dataset you're using for this assignment (e.g., means to access data, context, sample, variables, etc...)

   + *describe your data.*
   
- Third, create any animated plot with the `Play` button.  

```{r}
# add your codes 
```

- Fourth, describe your findings from the animated plot. 

